Clinical Trial Design Using A Stopped Negative Binomial Distribution

### 2018-02-12 sii/deveaux-kane-zelterman.tex
SII accepted the paper.

\begin{aug}
\author{\fnms{Michelle} \snm{DeVeaux} \ead[label=e1]{michelle.deveaux@yale.edu}}
\address{Department of Biostatistics\\ School of Epidemiology and Public Health \\ Yale University, New Haven, CT \\ \printead{e1}}
\and
\author{\fnms{Michael J.} \snm{Kane}\thanksref{t2}\corref{} \ead[label=e2]{michael.kane@yale.edu}}
\address{Department of Biostatistics\\ School of Epidemiology and Public Health \\ Yale University, New Haven, CT \\ \printead{e2}}
\and
\author{\fnms{Daniel} \snm{Zelterman} \ead[label=e3]{daniel.zelterman@yale.edu}}
\address{Department of Biostatistics\\ School of Epidemiology and Public Health \\ Yale University, New Haven, CT \\ \printead{e3}}
%\address{???\printead{e2}}

@@

\begin{abstract}
We introduce a discrete distribution suggested by curtailed
sampling rules common in early-stage clinical trials unifying the
negative, minimum negative (riff-shuffle), and
maximum negative binomial distributions. We derive the
distribution of the smallest number of independent and identically
distributed Bernoulli trials needed to observe either $s$ successes 
or $t$ failures. This report provides a closed-form expression for the 
mass function, moment generating function, and provides connections to other,
standard distributions.
\end{abstract}

@@

If all 17 patients are enrolled at once, as in the classic
design, then the sample size is 17. However, in most clinical trials the
patients are enrolled sequentially over time.
In the present example, observing seven
patients responding to treatment ends the trial and so the number of 
enrollees required could be as small as seven. Similarly 11
observed treatment non-responses would also end the trial. This sampling mechanism, in
which the experiment ends as soon as either predefined endpoint is reached, is
called {\em curtailed sampling}. Under curtailed sampling, the range of the
sample size for this trial is seven through 17.

\begin{figure*}
\centering
\includegraphics[width=0.8\textwidth]{snb-first-plot.pdf}
\caption{
The distribution of the sample size in a trial that stops after either seven patients
respond to treatment or 11 do not when $p=0.2$.
}
\label{fig:kane_viz_1}
\end{figure*}

\begin{figure*}[htbp!]
\centering
\includegraphics[width=\textwidth]{mean-and-variance.pdf}
\caption{
The mean and variance of the sample size for the probability of success $p$ 
when the trial stops after $s=7$ patients respond or $t=11$ fail to respond.
}
\label{fig:kane_viz_2}
\end{figure*}

The distribution of the number of trial enrollments is shown in 
Figure~\ref{fig:kane_viz_1}. There is relatively little probability mass
for values of sample sizes from seven through 10 since $p$ is small and it 
is unlikely the treatment will succeed quickly.
Figure~\ref{fig:kane_viz_2} shows the expected value and variance for the
number of trial enrollments varying the success rate $p$ between zero and one. When $p$ is small then the treatment
is likely to fail shortly after the $11^{\text{th}}$ enrollment.
When $p$ is large then the treatment is more likely to succeed and the 
number of enrollees approaches seven from above. 
When $p=0$ or $1$ then the sequence of resposes is deterministic and the 
variance is zero.
Values of $p$ between zero and one change the mixing proportions of 
the two endpoints. When $p$ is close to zero, most of the variance
is contributed from the failure endpoint. The saddle around $p=0.25$ results 
from a trade-off between the success and failure endpoints.

In the rest of this work, we derive the distribution of the number of 
enrollees needed
to observe either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB). 
In Section 2 we derive the distribution function and explores its properties.
Section 3 derives the moment generating function.
Section 4 describes the likelihood function.
Section 5 relates the distribution to other standard
distributions and connects the SNB tail probability to the binomial tail 
probability. Section 6 considers single arm, early-phase trial design with the 
SNB.

@@

Let $s$ and $t$ be positive integers.  Define the SNB random
variable $Y$ as the smallest
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either}
$\,s\,$ responders {\em or} $\,t\,$ non-responders. That is, the SNB 
distribution of $Y$ is the smallest integer such that either
$\sum_i^Y b_i = s$ or $\sum_i^Y 1-b_i = t$.

The sequence of Bernoulli random variables can be modeled as the process 
$\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ \mathbbm{1}_{\{ k-t < X_k < s\}}.
\end{equation*}
At each step a patient's outcome is measured. The process continues until either 
$X_k = s$ or $X_k = k-t$ corresponding to the success and failure boundaries.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{KanePlot.pdf}
\caption{
A hypothetical realization of the trial.
}
\label{fig:kane_viz}
\end{figure*}

@@

We next derive the mass function of $Y$. The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}.
\end{equation*}
In Proposition \ref{stopping_time} below, we show the probability mass function of $Y$ is
\begin{align} \label{eqn:pmf}
\mathbb{P} [Y=k] =& S(k, p, s) \ \mathbbm{1}_{\{s \leq k \leq s+t-1\}} \ + \nonumber \\
  & S(k, 1-p, t) \ \mathbbm{1}_{\{ t \leq k \leq s+t-1 \}}
\end{align}
where $\mathbbm{1}_{\{f\}}$ is the {\em indicator function}, taking the value 
of one if $f$ is true and zero otherwise, and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is a translated version of the negative binomial probability mass.

@@

The endpoint $X_k = s$ can only occur if $X_{k-1} = s-1$ followed by a success. That is, 
\begin{align} \label{prop11}
\mathbb{P} [X_k =s ] \;&=\; p \, \mathbb{P} [X_{k-1} = s-1 ] \nonumber \\
 &=\; p\, {k-1 \choose s-1} p^{s-1} (1-p)^{k-s} \nonumber \\
 &=\; {k-1 \choose s-1} p^{s} (1-p)^{k-s}.
\end{align}
This expression is given
in (\ref{eqn:N}). 
Similarly, the probability a given realization reaches the endpoint $X_k = k-t$
satisfies 
\begin{align} \label{prop12}
\mathbb{P} [X_k = k-t ] \;&=\; (1-p) \, \mathbb{P} [X_{k-1} = k-t ] \nonumber \\
 &=\; (1-p)\, {k-1 \choose t-1} (1-p)^{t-1} p^{k-t} \nonumber \\
&=\; {k-1 \choose t-1} (1-p)^{t} p^{k-t}.
\end{align}
The result follows by summing (\ref{prop11}) and (\ref{prop12}).

To show (\ref{eqn:pmf}) sums to one, define
\begin{equation*} 
R = \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t).
\end{equation*}
Substitute $i=k-s$ in the first summation and $j=k-t$ in the second. Then
$R$ can be written as the cumulative distribution function (CDF) of two
negative binomial distributions:
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1
\end{align*}
using (\ref{ribf}). This completes the proof (\ref{eqn:pmf}) is the distribution 
of the stopping time and is a valid probability mass function.
\end{proof}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.8\textwidth]{likelihood.pdf}
\end{center}
\caption{Shapes of the likelihood function for given values of $Y$ with 
$s=7$ and $t=11$.
\label{likelihood.fig}}
\end{figure*}

@@

\begin{prop} Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{align} \label{eqn:mgf}
\mathbb{E}~e^{xY} = &\left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) \ + \nonumber \\
  & \left(\frac{qe^x}{1-pe^x}\right)^t \mathcal{I}_{1-pe^x}(t, s)
\end{align}
for $q = 1-p$ and is defined for 
$x < \min \left\{\log(1/p), \log(1/q) \right\}$.
\end{prop}
The moment generating function for the SNB is calculated in a manner similar to 
that of two negative binomial distributions. Appendix 1 provides a proof for the derivation.

@@

Let $Y$ be distributed SNB($p, s, t$). Then, the likelihood of $Y$ is proportional to a mixture of Beta distributions.
\begin{equation*} \label{eqn:likelihood}
\mathcal{L} (p | s, t, Y) \propto B_1 \, \mathbbm{1}_{\{s \leq Y_1\}} + B_2 \, \mathbbm{1}_{\{t \leq Y_1 \}}
\end{equation*}
where $B_1 =$ Beta$\left(s, Y - s\right)$ and $B_2 =$ Beta$\left(Y - t, t\right)$ and
\begin{equation*}
\text{Beta}\left( \alpha, \beta \right) = \frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\, \Gamma(\beta)} 
p^{\alpha-1} (1-p)^{\beta-1}.
\end{equation*}

@@

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{shapes.pdf}
\end{center}
\caption{Different shapes of the SNB mass function with parameters 
($p$, $s$, $t$) given. Black areas indicate the mass contributed by reaching
$s$ responders before $t$ non-responders. Grey indicates
mass contributed by reaching $t$ non-responders first. \label{shapes.fig}}
\end{figure*}

@@

\begin{figure*}
\centering
\subfloat[SNB distribution]{\includegraphics[width=0.5\textwidth]{snb_density.pdf}}
\hfill
\subfloat[binomial distribution]{\includegraphics[width=0.5\textwidth]{bin_density.pdf}}
\caption{
SNB(0.2, 7, 11) with mass contributed from 
7 responders (black) or 11 non-responders (grey) along with 
Binomial(0.2, 17) with at least 2 responders (black) or fewer (grey).
}
\label{fig:snb_bin_compare}
\end{figure*}

@@

\begin{figure*}
\centering
\subfloat[ROC Curve]{\includegraphics[width=0.45\textwidth]{power-vs-significance.pdf}}
\hfill
\subfloat[Expected Sample Size]{\includegraphics[width=0.45\textwidth]{expected-sample-size.pdf}}
\caption{
The ROC and expected sample sizes of all trial designs where the maximum
number of patients is 17, $p=0.2$ under the null, and $p=0.4$ under
the alternative. The numerical values in the ROC curve indicate the number
of responses $s$ required to reach the success endpoint.
}
\label{fig:all_sample_sizes}
\end{figure*}

Consider the problem of designing a trial using curtailed sampling and the SNB. 
Assume that the maximum number of patients $n$ is given along with the null 
and alternative hypotheses. In this case, the parameter $s$ tells how many 
responses are needed to reach the success endpoint, can vary between 1 and 
$n-1$. For fixed $s$ and $n$  the value of $t$ is determined by the relation 
$t=n-s+1$. The ROC curves for all trial designs in the prototype, 
where $n=17$, $p=0.2$ under the null, and $p=0.4$ under the alternative, are 
shown in Figure~\ref{fig:all_sample_sizes} (a). In designing these trials, 
small significance and large power values are attained by either increasing 
the value of $n$ or increasing the difference between the alternative and null 
response rates. Figure~\ref{fig:all_sample_sizes} (b) shows the expected 
number of enrollees for each of these trials as a function of the number of 
responses required to reach the success endpoint.
The curve reaches its maximum of 15 patients when $s=5$. The prototype 
design ($p=0.2$, $s=7$, and $t=11$) has an expected sample size of 14 enrollees.

\section{Appendix: Proof of the MGF of the SNB}
Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{equation} 
\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s)
\end{equation}
for $q = 1-p$ and is defined for 
$x < \min \left\{\log(1/p), \log(1/q) \right\}$.
\begin{proof}
The MGF of the SNB is, by definition:
\begin{equation*}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose t-1} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{align} \label{eqn:first_sum}
\mathbb{E}~e^{xY} =& 
  \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe^x)^{s} (qe^x)^{k-s} \ + \nonumber \\
  & \sum_{k=t}^{s+t-1}{k-1 \choose t-1} (qe^x)^t (pe^x)^{k-t}.
\end{align}
The first summation in (\ref{eqn:first_sum}) satisfies
\begin{align*}
\sum_{k=s}^{s+t-1} & {k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} \\ 
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{1-qe^x}(s, t).
\end{align*}
Since the $p$ parameter in $\mathcal{I}_p$ has domain zero 
to one, we have $0 \leq pe^x < 1$. This implies $x < -\log(p)$.
A similar expression can be derived for the second summation in 
(\ref{eqn:first_sum}) and results in
the constraint $x < -\log(1-p)$.
\end{proof}

\section*{Acknowledgements}
This research was supported by grants R01CA131301, 
R01CA157749, R01CA148996, R01CA168733, and PC50CA196530 awarded by the 
National Cancer Institute along with support from the Yale Comprehensive Cancer 
Center and the Yale Center for Outcomes Research. We would also like 
to thank Rick Landin at Ignyta Inc. for his suggestions.

@@

>>>>

\begin{aug}
\author{\fnms{Michelle} \snm{DeVeaux} \ead[label=e1]{michelle.deveaux@regeneron.com}}
\address{Regeneron Pharmaceuticals Inc.\\ Tarrytown, NY\\ \printead{e1}}
\and
\author{\fnms{Michael J.} \snm{Kane}\thanksref{t2}\corref{} \ead[label=e2]{michael.kane@yale.edu}}
\address{Department of Biostatistics\\ School of Epidemiology and Public Health \\ Yale University, New Haven, CT \\ \printead{e2}}
\and
\author{\fnms{Daniel} \snm{Zelterman} \ead[label=e3]{daniel.zelterman@yale.edu}}
\address{Department of Biostatistics\\ School of Epidemiology and Public Health \\ Yale University, New Haven, CT \\ \printead{e3}}
%\address{???\printead{e2}}

@@

\begin{abstract}
We introduce a discrete distribution suggested by curtailed
sampling rules common in early-stage clinical trials. We derive the
distribution of the smallest number of independent and identically
distributed Bernoulli trials needed to observe either $s$ successes 
or $t$ failures. This report provides a closed-form expression for the 
mass function, moment generating function, and provides connections to other,
standard distributions.
\end{abstract}

@@

If all 17 patients are enrolled at once, as in the classic
design, then the sample size is 17. However, in most clinical trials the
patients are enrolled sequentially over time.
In the present example, observing seven
successful patients ends the trial and so the number of enrollees required
could be as small as seven. Similarly 11
observed treatment failures would also end the trial. This sampling mechanism, in
which the experiment ends as soon as either predefined endpoint is reached, is
called {\em curtailed sampling}. Under curtailed sampling, the range of the
sample size for this trial is seven through 17.

The distribution of the number of trial enrollments is shown in 
Figure~\ref{fig:kane_viz_1}. There is relatively little probability mass
for values of sample sizes from seven through 10 since $p$ is small and it 
is unlikely the treatment will succeed quickly.
Figure~\ref{fig:kane_viz_2} shows the expected value and variance for the
number of trial enrollments varying the success rate $p$ between zero and one. When $p$ is small then the treatment
is likely to fail shortly after the $11^{\text{th}}$ enrollment.
When $p$ is large then the treatment is more likely to succeed and the 
number of enrollees approaches seven from above.

When $p=0$ or $1$ then the processes are deterministic and variance is zero.
Values of $p$ between zero and one change the mixing proportions of 
the two endpoints. When $p$ is close to zero, most of the variance
is contributed from the failure endpoint. The saddle around $p=0.25$ results 
from a trade-off between the success and failure endpoints.

In the rest of this work, we derive the distribution of the number of 
enrollees needed
to observe either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB). 
In Section 2 we derive the distribution function and explores its properties.
Section 3 derives the moment generating function.
Section 4 describes the likelihood function.
Section 5 describes the posterior and predictive probability distributions. Section 6 relates the distribution to other standard
distributions and connects the SNB tail probability to the binomial tail 
probability. Section 7 shows how to design a trial using the SNB in a prototypical setting.

@@

Let $s$ and $t$ be positive integers.  Define the SNB random
variable $Y$ as the smallest
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either}
$\,s\,$ responders {\em or} $\,t\,$ non-responders. That is, the SNB 
distribution of $Y$ is the smallest integer such that either
$\sum_i^Y b_i = s$ or $\sum_i^Y (1-b_i) = t$.

The sequence of Bernoulli random variables can be modeled as the process 
$\mathbf{X} = \left\{X_k : k = 0,1,... \right\}$
with $X_0=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ \mathbbm{1}_{\{ k-t < X_k < s\}}.
\end{equation*}
where $\mathbbm{1}_{\{f\}}$ is the {\em indicator function}, taking the value 
of one if $f$ is true and zero otherwise, and
At each step a patient's outcome is measured. If it is a response, the response count increases, otherwise it stays the same. The process continues until either 
$X_k = s$ or $X_k = k-t$ corresponding to the success and failure boundaries

@@

We next derive the probability mass function of $Y$. The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}.
\end{equation*}
In Proposition \ref{stopping_time} below, we show the probability mass function of $Y$ is
\begin{align} \label{eqn:pmf}
\mathbb{P} [Y=k] =& \ S(k, p, s) \ \mathbbm{1}_{\{s \leq k \leq s+t-1\}} \nonumber \\
  & + S(k, 1-p, t) \ \mathbbm{1}_{\{ t \leq k \leq s+t-1 \}}
\end{align}
where
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is a translated version of the negative binomial probability mass.

@@

The endpoint $X_k = s$ can only occur if $X_{k-1} = s-1$ followed by a success. That is, 
\begin{align} \label{prop11}
\mathbb{P} [X_k =s ] \;&=\; p \, \mathbb{P} [X_{k-1} = s-1 ] \nonumber \\
 &=\; p\, {k-1 \choose s-1} p^{s-1} (1-p)^{k-s} \nonumber \\
 &=\; {k-1 \choose s-1} p^{s} (1-p)^{k-s}.
\end{align}
This expression is given
in (\ref{eqn:N}). 
Similarly, the probability a given realization reaches the endpoint $X_k = k-t$
satisfies 
\begin{align} \label{prop12}
\mathbb{P} [X_k = k-t ] \;&=\; (1-p) \, \mathbb{P} [X_{k-1} = k-t ] \nonumber \\
 &=\; (1-p)\, {k-1 \choose t-1} (1-p)^{t-1} p^{k-t} \nonumber \\
&=\; {k-1 \choose t-1} (1-p)^{t} p^{k-t}.
\end{align}
The result (\ref{eqn:pmf}) follows by summing (\ref{prop11}) and (\ref{prop12}).

To show (\ref{eqn:pmf}) sums to one, define
\begin{equation*} 
R = \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t).
\end{equation*}
Substitute $i=k-s$ in the first summation and $j=k-t$ in the second. Then
$R$ can be written as the cumulative distribution function (CDF) of two
negative binomial distributions:
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1
\end{align*}
using (\ref{ribf}). This completes the proof that (\ref{eqn:pmf}) is the distribution 
of the stopping time and it is a valid probability mass function.
\end{proof}

@@

@@

\begin{prop} Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s)
\end{equation}
for $q = 1-p$ and is defined for 
$x < \min \left\{\log(1/p), \log(1/q) \right\}$.
\end{prop}
The moment generating function for the SNB is calculated in a manner similar to 
that of two negative binomial distributions. Appendix 1 provides a proof for the derivation.

@@

In this section we derive the likelihood function for the SNB for a single trial. In early-stage clinical trial only a single trial is performed, usually because of resource constraints, and the object of interest is $p$ parameter, which determines if a therapy will be delivered to the market. A multi-sample version is less relevant for this use case but is represented as a product of mixtures of Beta distributions. Deriving it's theoretical characteristics not straight-forward.

Let $Y$ be distributed SNB($p, s, t$). Then, the likelihood of $Y$ is proportional to a mixture of Beta distributions.
\begin{equation*} \label{eqn:likelihood}
\mathcal{L} (p | s, t, Y) = B_1 \, \mathbbm{1}_{\{s \leq Y_1\}} + B_2 \, \mathbbm{1}_{\{t \leq Y_1 \}}
\end{equation*}
where $B_1 =$ Beta$\left(s, Y - s\right)$ and $B_2 =$ Beta$\left(Y - t, t\right)$ and
\begin{equation*}
\text{Beta}\left( \alpha, \beta \right) = \frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\, \Gamma(\beta)} 
p^{\alpha-1} (1-p)^{\beta-1}.
\end{equation*}

@@

\section{The Posterior and Predictive Probability Distribution} \label{bayesian_extension}

Consider the Bayesian setting where $Y$ is an SNB($P$, $s$, $t$) distribution and the rate parameter, $P$ 
is distributed as Beta($\alpha$, $\beta$). The posterior distribution $P|Y$ is proportional to the 
likelihood, given by the function
%f_P(p, \alpha, \beta ) f_{Y|P}(p, k, s, t) &= 
\begin{align} \label{eqn:ptl}
f_{P|Y}(p, k, s, t, \alpha, \beta) \; \propto \ &
  \ \frac{ {k-1 \choose s-1} }{B(\alpha, \beta)} p^{\alpha +s -1} 
    (1-p)^{k+\beta-s-1} \\
  & + \frac{ {k-1 \choose t-1} }{B(\alpha, \beta)} p^{k+\alpha -t -1} 
    (1-p)^{\beta+t-1} \nonumber
\end{align}
where $0 \leq p \leq 1$ and $\min(s, t) \leq k \leq s+k-1$.

The predictive distribution of the SNB can be found as by integrating
$p$ over the interval zero to one and applying the definition of the 
beta function.
\begin{align} \label{eqn:predictive}
f_{Y}(&k, s, t, \alpha, \beta) = 
  \int_0^1 f_P(p | \alpha, \beta )  f_{Y|P}(p, k, s, t) dp \\ 
  = & {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
     \nonumber \\
  & + {k-1 \choose t-1} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} \nonumber
\end{align}
If both $\alpha$ and $\beta$ are non-negative integers then the predictive
distribution is a mixture of negative-hypergeometric distributions. 
\begin{align*} \label{eqn:hypergeo}
f_{Y}(k, s, t, \alpha, \beta) = & 
  \frac{ {k - 1 \choose s - 1 } 
         {\alpha + \beta \choose \alpha} }{
         {\alpha + \beta + k - 1 \choose \alpha + s}}
  \frac{\alpha}{\alpha + \beta}
  \frac{ \beta }{ k-s+\beta } \\ & +
  \frac{ {k - 1 \choose t - 1} 
         {\alpha + \beta \choose \beta} 
         }{
         {\alpha + \beta + k -1 \choose t + \beta}
       }
  \frac{ \beta }{ \alpha + \beta}
  \frac{ \alpha }{ k-t + \alpha}
\end{align*}
The ratio of combinations in the first term can be interpreted as the
probability of $s-1$ responders from $k-1$ patients in $\alpha + s$ draws
from a population size of $\alpha + \beta + k - 1$. This value is multiplied
by $\alpha / (\alpha + \beta)$, the expected response rate of the prior. 
The final term in the product weights the prior based on the number
of non-responders ($k-s$). Terms in the second summand are interpreted similarly
for non-responders.

The ratio of (\ref{eqn:ptl}) divided by (\ref{eqn:predictive}) gives the 
posterior distribution of $P$. It is a mixture of beta distributions. The
mixing parameter depend on the endpoints ($s$ and $t$), the number of enrollees needed to reach an endpoint ($k$), and the prior parameters ($\alpha$ and
$\beta$).

@@

@@

@@

Consider the problem of designing a trial using curtailed sampling and the SNB. Assume that the maximum number of patients $n$ is given along with the null and alternative hypotheses. In this case, the parameter $s$ tells how many responses are needed to reach the success endpoint, can vary between 1 and $n-1$. For fixed $s$ and $n$  the value of $t$ is determined by the relation $t=n-s+1$. The ROC curves for all trial designs in the prototype, where $n=17$, $p=0.2$ under the null, and $p=0.4$ under the alternative, are shown in Figure~\ref{fig:all_sample_sizes_power_vs_signif}. In designing these trials, small significance and large power values are attained by either increasing the value of $n$ or increasing the difference between the alternative and null response rates. Figure~\ref{fig:all_sample_sizes} shows the expected number of enrollees for each of these trials as a function of the number of responses required to reach the success endpoint.

The expected number of enrollees for each trial is shown in Figure~\ref{fig:all_sample_sizes}. The curve reaches its maximum of 15 patients when $s=5$. The prototype design ($p=0.2$, $s=7$, and $t=11$) has an expected sample size of 14 enrollees.

\section*{Acknowledgements}
This research was supported by grants R01CA131301, 
R01CA157749, R01CA148996, R01CA168733, and PC50CA196530 awarded by the 
National Cancer Institute along with support from the Yale Comprehensive Cancer 
Center and the Yale Center for Outcomes Research. We would also like 
to thank Rick Landin at LaJolla Pharmaceutical for his suggestions.

\section*{APPENDIX: PROOF OF PROPOSITION 2}
%Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
%Then the moment generating function (MGF) of $Y$ is
%\begin{equation} 
%\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
%  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
%  \mathcal{I}_{1-pe^x}(t, s)
%\end{equation}
%for $q = 1-p$ and is defined for 
%$x < \min \left\{\log(1/p), \log(1/q) \right\}$.
%\begin{proof}
The MGF of the SNB is, by definition:
\begin{align*}
\mathbb{E}~e^{xY} = & \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} \\
  & + \sum_{k=t}^{s+t-1} {k-1 \choose t-1} p^{k-t} q^t e^{kx}
\end{align*}
and can be rewritten as:
\begin{align} \label{eqn:first_sum}
\mathbb{E}~e^{xY} = & \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe^x)^{s} (qe^x)^{k-s} \nonumber \\
  & + \sum_{k=t}^{s+t-1}{k-1 \choose t-1} (qe^x)^t (pe^x)^{k-t}.
\end{align}
The first summation in (\ref{eqn:first_sum}) satisfies
\begin{align*}
\sum_{k=s}^{s+t-1} &{k-1 \choose s-1}  (pe)^{sx} (qe^x)^{k-s}  \\
  & = \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s \\
  & = \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{1-qe^x}(s, t).
\end{align*}
Since the $p$ parameter in $\mathcal{I}_p$ has domain zero 
to one, we have $0 \leq pe^x < 1$. This implies $x < -\log(p)$.
A similar expression can be derived for the second summation in 
(\ref{eqn:first_sum}) and results in
the constraint $x < -\log(1-p)$.
%\end{proof}

@@

\pagebreak

\onecolumn

\section*{Figures}

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{snb-first-plot.pdf}
\caption{
The distribution of the sample size in a trial that stops after either seven patients
respond to treatment or 11 do not when $p=0.2$.
}
\label{fig:kane_viz_1}
\end{figure}

\pagebreak

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{mean-and-variance.pdf}
\caption{
The mean and variance of the sample size for the probability of success $p$ 
when the trial stops after $s=7$ patients respond or $t=11$ fail to respond.
}
\label{fig:kane_viz_2}
\end{figure}

\pagebreak

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{KanePlot.pdf}
\caption{
A hypothetical realization of the trial.
}
\label{fig:kane_viz}
\end{figure}

\pagebreak

\begin{figure}[htbp!]
\begin{center}
\includegraphics[width=\textwidth]{likelihood.pdf}
\end{center}
\caption{Shapes of the likelihood function for given values of $Y$ with 
$s=7$ and $t=11$.
\label{likelihood.fig}}
\end{figure}

\pagebreak

\begin{figure}[htbp!]
\begin{center}
\includegraphics[width=0.8\textwidth]{shapes.pdf}
\end{center}
\caption{Different shapes of the SNB mass function with parameters 
($p$, $s$, $t$) given. Black areas indicate the mass contributed by reaching
$s$ responders before $t$ non-responders. Grey indicates
mass contributed by reaching $t$ non-responders first. \label{shapes.fig}}
\end{figure}

\pagebreak

\begin{figure}[htbp!]
\centering
\subfloat[SNB distribution]{\includegraphics[width=0.5\textwidth]{snb_density.pdf}}
\hfill
\subfloat[binomial distribution]{\includegraphics[width=0.5\textwidth]{bin_density.pdf}}
\caption{
SNB(0.2, 7, 11) with mass contributed from 
7 responders (black) or 11 non-responders (grey) along with 
Binomial(0.2, 17) with at least 2 responders (black) or fewer (grey).
}
\label{fig:snb_bin_compare}
\end{figure}

\pagebreak

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{power-vs-significance.pdf}
\caption{
ROC curve of the all trial designs where the maximum number of patients is 17, $p=0.2$ under the null and $p=0.4$ under the alternative. The numerical values indicate the number of responses $s$ required reach the success endpoint.
}
\label{fig:all_sample_sizes_power_vs_signif}
\end{figure}

\pagebreak

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{expected-sample-size.pdf}
\caption{
The expected sample size all trial designs where the maximum number
of patients is 17 and $p=0.2$.}
\label{fig:all_sample_sizes}
\end{figure}

\pagebreak
